---
title: "DisturbanceChapter_GitHub"
output: html_document
date: "2026-04-06"
---
Summarized Code/Clean Code: 

Community Composition:

loading libraries
```{r}
library(stringr)
library(dplyr)
library(stats)
library(ggplot2)
library(ggpubr)
library(vegan)
library(cowplot)
library(tidyverse)
library(phyloseq)
library(microbiome)
library("funfuns")
library(pairwiseAdonis)
library(viridis)
library("microViz")
```

libraries:
```{r}
library(ggplot2)
library(cowplot)
library(phyloseq)
library(car)
library(ggpubr)
library(vegan) #vegan 2.6-4
library(dada2)
library(RColorBrewer)
library(dplyr)
library(tidyverse)
```

```{r}
getwd()
```

read in clean phyloseq object & seqtab object from dada2pipeline with rel abundance counts applied
(ps.cleaner.rel)

```{r}
# load it back into R
ps.cleaner.clr <- readRDS("ch4.ps.cleaner.clr.rds")
```

applying a CLR (Centered Log-Ratio) transformation 
making a sequence table (already loaded in clr version above, so skip this for now)
```{r}
ps.cleaner.clr <- microbiome::transform(ps.cleaner.rel,'clr')
seqtab.clr <- data.frame(otu_table(ps.cleaner.clr))
```

so....
- ps.cleaner.rel --> relative abundances of sample counts
- ps.cleaner.clr --> relative abundances with the clr transformation 
- seqtab.clr --> clr transformed sequence table

subsetting into 3 experiments: 
```{r}
ps.summer <- subset_samples(ps.cleaner.clr, Experiment == "Summer")
ps.field <- subset_samples(ps.cleaner.clr, Experiment == "Field")
ps.temp <- subset_samples(ps.cleaner.clr, Experiment == "Temp")
```

PCOA plots to visualize spread before stats!
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
~Lab Dist (Summer) FIGURE 1A
```{r}
# Make Treatment a factor in the desired order
ps.summer@sam_data$Treatment <- factor(
  ps.summer@sam_data$Treatment,
  levels = c("ND", "Few Small", "Many Small", "Few Large", "HD")
)

# Now plot
sum.treat.temp.week.ord <- plot_ordination(
  ps.summer,
  ordinate(ps.summer, "PCoA", "euclidean")
) +
  geom_point(
    aes(
      fill = Treatment,
      shape = as.factor(Temperature)
    ),
    size = 4,
    color = "black",
    alpha = 0.8
  ) +
  stat_ellipse(
    aes(group = as.factor(Week), color = as.factor(Week)),
    linetype = "dashed",
    size = 1
  ) +
  scale_color_manual(
    name = "Week",
    values = c("1" = "tan4", "3" = "orange", "5" = "brown")
  ) +
  scale_fill_manual(
    name = "Treatment",
    values = c(
      "ND" = "#999999",
      "Few Small" = "#d95f02",
      "Many Small" = "#0072B2",
      "Few Large" = "#1b9e77",
      "HD" = "purple3"
    )
  ) +
  scale_shape_manual(
    name = "Temperature (°C)",
    values = c("22" = 21, "26" = 22, "30" = 24)
  ) +
  guides(
    fill = guide_legend(
      override.aes = list(shape = 21),
      keywidth = 0.4, keyheight = 0.4
    ),
    shape = guide_legend(
      override.aes = list(fill = "white"),
      keywidth = 0.4, keyheight = 0.4
    )
  ) +
  theme_classic(base_size = 22) +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.spacing.y = unit(0.0005, "cm"),
    legend.key.size = unit(0.4, "lines")
  )

sum.treat.temp.week.ord
```
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
~Field Dist Experiment(Field) Figure 3A
```{r}
# Make Treatment a factor
sample_data(ps.field)$Treatment <- factor(
  sample_data(ps.field)$Treatment,
  levels = c("ND", "Few Small", "Many Small", "HD")
)

# Ordination
ord <- ordinate(ps.field, method = "PCoA", distance = "euclidean")

# Plot
field.treat.ord <- plot_ordination(ps.field, ord) +
  geom_point(
    aes(fill = Treatment),
    size = 4,
    shape = 21,
    color = "black",
    alpha = 0.8
  ) +
  stat_ellipse(
   aes(group = Treatment, color = Treatment),
  linetype = "dashed",
   size = 1
  ) +
  theme_classic(base_size = 22) +
  scale_fill_manual(
    name = "Treatment",
    values = c(
      "ND" = "#999999",
      "Few Small" = "#d95f02",
      "Many Small" = "#0072B2",
      "HD" = "purple3"
    )
  ) +
  
  scale_color_manual(
    name = "Treatment",
    values = c(
"ND" = "#999999",
      "Few Small" = "#d95f02",
      "Many Small" = "#0072B2",
      "HD" = "purple3")
  ) +
  
  guides(
    fill = guide_legend(override.aes = list(shape = 21)),
    color = "none"
  ) +
  theme(
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16)
  )

field.treat.ord
```
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
~ Temperature Experiment (Temp) Figure 4A

```{r}
# Make factors
sample_data(ps.temp)$Week <- factor(sample_data(ps.temp)$Week, levels = c("1", "3", "5"))
sample_data(ps.temp)$Temperature <- factor(sample_data(ps.temp)$Temperature, levels = c("26", "29", "31", "33"))

# Ordination
ord <- ordinate(ps.temp, method = "PCoA", distance = "euclidean")

# Plot
temp.temp.week.ord <- plot_ordination(ps.temp, ord) +
  geom_point(
    aes(color = Temperature, shape = Week),
    size = 4,
    alpha = 0.8
  ) +
  stat_ellipse(
    aes(group = Week, color = Week),  # Ellipse by Week
    linetype = "dashed",
    size = 1
  ) +
  scale_color_manual(
    values = c(
      "26" = "blue3", "29" = "darkgreen", "31" = "magenta4", "33" = "red3",
      "1" = "tan4", "3" = "orange", "5" = "brown"  # for Week ellipses
    )
  ) +
  guides(
    shape = guide_legend(override.aes = list(color = "black")), # Week shapes
    color = guide_legend(override.aes = list(shape = 16))        # Temp colors
  ) +
  theme_classic(base_size = 22) +
  theme(
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 18),
    legend.text = element_text(size = 18),
    legend.title = element_text(size = 18)
  )

temp.temp.week.ord
#good enough, will edit key later
```
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
Set up for Beta Dispersion ~ ALL EXPERIMENTS

-subset seqtab by experiment (sequence table)
```{r}
sum.seqtab <- readRDS("chap4sumseqtab.clr.rds")
field.seqtab <- readRDS("chap4fieldseqtab.clr.rds")
temp.seqtab <- readRDS("chap4tempseqtab.clr.rds")
```

-also samdf is all the metadata, this needs to be read it and subset by experiment as well (already subsetted, so just reading in) (metadata)
```{r}
sum.samdf <- read.csv("chap4sum.samdf.csv")
field.samdf <- read.csv("chap4field.samdf.csv")
temp.samdf <- read.csv("chap4temp.samdf.csv")
```

Make a Distance Matrix: veg-distance calculates pairwise distances between samples (using euclidean distances)
```{r}
sum.dist.all <- vegdist(sum.seqtab, method = "euclidean")
field.dist.all <- vegdist(field.seqtab, method = "euclidean")
temp.dist.all <- vegdist(temp.seqtab, method = "euclidean")
```
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
SUMMER (OR LAB DIST) Beta Dispersion: 
*looks at spread of points in a multidemensional space
*is beta dispersion differs a lot, PERMANOVA results may be biased

```{r}
sum.bet.all1 <- betadisper(sum.dist.all,sum.samdf$Temperature)
sum.bet.all2 <- betadisper(sum.dist.all,sum.samdf$Week)
sum.bet.all3 <- betadisper(sum.dist.all,sum.samdf$Treatment)
```

```{r}
#temp
sum.bet.all1
anova(sum.bet.all1)
```
*no differences in dispersion between temps

```{r}
#week
sum.bet.all2
anova(sum.bet.all2)
TukeyHSD(sum.bet.all2)
```
*diff between:
- weeks 5 & 3
- weeks 5 & 1 


```{r}
#treatment
sum.bet.all3
anova(sum.bet.all3)
```
*no differences in dispersion between treatments
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
FIELD Beta Dispersion: 

```{r}
field.bet.all <- betadisper(field.dist.all,field.samdf$Treatment)
field.bet.all
anova(field.bet.all)
```
*no differences in dispersion between  treatments
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
TEMP Beta Dispersion: 

```{r}
#temp
temp.bet.all1 <- betadisper(temp.dist.all,temp.samdf$Temperature)
anova(temp.bet.all1)
```
*no differences in dispersion betweentemperatures


```{r}
#week
temp.bet.all2 <- betadisper(temp.dist.all,temp.samdf$Week)
anova(temp.bet.all2)
TukeyHSD(temp.bet.all2)
```
*diff between:
- weeks 3 & 1
- weeks 5 & 1 

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
LAB (SUMMER) RESULTS! 

Beta Diversity & Comm Comp Stats
~ full permanova model
```{r}
sum.samdf$Temperature <- as.factor(sum.samdf$Temperature)
sum.samdf$Week <- as.factor(sum.samdf$Week)
sum.samdf$Treatment <- as.factor(sum.samdf$Treatment)
adonis2(sum.dist.all ~ Treatment * Temperature *Week , data=sum.samdf, permutations=9999)
```
Treatment                   4    90414 0.08753 1.7698 0.0001 ***
Temperature                 1    23652 0.02290 1.8519 0.0021 ** 
Week                        1    59115 0.05723 4.6286 0.0001 ***
Treatment:Temperature       3    53248 0.05155 1.3897 0.0026 ** 

Since treatment:temp interaction, need to tear apart interaction 
week has a significant difference, may be influence by beta dispersion 

subset data ~ are treatments different at 22C only? YES
Example:
```{r}
# Logical index for samples at Temperature 22
temp22_idx <- sum.samdf$Temperature == "22"

# Subset the distance matrix (use square brackets once)
dist_temp22 <- as.dist(as.matrix(sum.dist.all)[temp22_idx, temp22_idx])

# Subset metadata to match
meta_temp22 <- sum.samdf[temp22_idx, ]

# Run PERMANOVA
adonis2(dist_temp22 ~ Treatment, data = meta_temp22, permutations = 9999)


```
subset data ~ are treatments different at 26C only? YES
```{r}
# Logical index for samples at Temperature 22
temp26_idx <- sum.samdf$Temperature == "26"

# Subset the distance matrix (use square brackets once)
dist_temp26 <- as.dist(as.matrix(sum.dist.all)[temp26_idx, temp26_idx])

# Subset metadata to match
meta_temp26 <- sum.samdf[temp26_idx, ]

# Run PERMANOVA
adonis2(dist_temp26 ~ Treatment, data = meta_temp26, permutations = 9999)
```
subset data ~ are temperatures different between the treatments? 
Yes, for ND & Few Small 
No, for Many Small & Few Large 
```{r}
# Get unique treatment levels
treatments <- unique(sum.samdf$Treatment)

# Loop over each treatment to test for temperature effect
for (treat in treatments) {
  
  cat("\n--- PERMANOVA: Temperature differences within Treatment =", treat, "---\n")
  
  # Logical index for samples matching the treatment
  treat_idx <- sum.samdf$Treatment == treat
  
  # Subset distance matrix and metadata
  dist_subset <- as.dist(as.matrix(sum.dist.all)[treat_idx, treat_idx])
  meta_subset <- sum.samdf[treat_idx, ]
  
  # Check if at least 2 temperature levels are present
  if (length(unique(meta_subset$Temperature)) > 1) {
    # Run PERMANOVA
    result <- adonis2(dist_subset ~ Temperature, data = meta_subset, permutations = 9999)
    print(result)
  } else {
    cat("Not enough temperature levels to compare for this treatment.\n")
  }
}

```

interaction plots 
```{r}
ordination_df2 <- read.csv("ordination_df2.csv")
```

figure 1B (left)
```{r}
#final interaction plot 1 for summer
ordination_df2$Treatment <- factor(
  ordination_df2$Treatment,
  levels = c("ND", "Few Small", "Many Small", "Few Large", "HD")
)
sum.intplot1 <- ggplot(ordination_df2, aes(
  x = as.factor(Temperature),
  y = PCoA1,
  color = Treatment,
  group = Treatment
)) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  scale_color_manual(values = c(
    "ND" = "#999999",        # neutral gray
    "Few Small" = "#d95f02", # orange
    "Many Small" = "#0072B2",# magenta-pink
    "Few Large" = "#1b9e77", # teal-green
    "HD" = "purple3"         # purple
  )) +
  labs(
    title = "",
    x = "Temperature",
    y = "PCoA Axis 1 score",
    color = "Treatment"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 14)
  )

sum.intplot1


```
figure 1B (right)
```{r}
# Then create the plot
sum.intplot2 <- ggplot(ordination_df2, aes(x = Treatment, y = PCoA1, color = as.factor(Temperature), group = Temperature)) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  scale_color_manual(
    values = c(
      "22" = "blue3",
      "26" = "red3"
    )
  ) +
  labs(
    title = "",
    x = "Grazing Treatment",
    y = "PCoA Axis 1 score",
    color = "Temperature"
  ) +
  
   theme_minimal(base_size = 16) +   # ← increases all font sizes (default = 11)
  theme(
    axis.title = element_text(size = 20),
    axis.text = element_text(size = 13),
    legend.title = element_text(size = 20),
    legend.text = element_text(size = 17)
  )


sum.intplot2
```

week alone (simple main effect)
```{r}
sum.samdf$Week <- as.factor(sum.samdf$Week)
adonis2(sum.dist.all ~ Week , data=sum.samdf, permutations=9999)
pairwise.adonis2(sum.dist.all ~ Week , data = sum.samdf, permutations = 9999)
```

Bar plots *UPDATED

upload mdf files, what are they?  a tidy data frame of relative abundances per sample and taxonomic group, formatted for plotting stacked bar plots ~ we want to group the stacked bar plots so we don't have a bunch of rare species, lets look
```{r}
mdffam.sum <- read.csv("ch4.mdffam.summer.csv")
mdffam.field <- read.csv("ch4.mdffam.field.csv")
mdffam.temp <- read.csv("ch4.mdffam.temp.csv")
```

calculates the total abundance of each Family across all samples, sorts them by the rel abundance, and converts them into percentages of the total 

groups rare families into one group (<1% rel abundance) 
groups taxa with "_" which means it was assigned at a higher taxonomy 
```{r}
# Aggregate total relative abundance per Family across all samples
top_fam1 <- aggregate(Abundance ~ Family, data = mdffam.sum, FUN = sum)

# Sort in descending order
top_fam1 <- top_fam1[order(-top_fam1$Abundance), ]

# Calculate the total abundance
total_abundance <- sum(top_fam1$Abundance)

# Calculate percentage for each Family
top_fam1$Percentage <- (top_fam1$Abundance / total_abundance) * 100

# Group rare families (<1%) and those lacking family-level assignment
top_fam1$Family_grouped <- ifelse(
  grepl("_", top_fam1$Family),             # condition 1: contains "_"
  "Assigned at higher taxonomy",           # assign this label
  ifelse(
    top_fam1$Percentage < 1,              # condition 2: <1% abundance
    "Rare Families",                       # assign this label
    top_fam1$Family                        # else keep original Family
  )
)

# View the result
print(top_fam1)
```
~~~~~~~~~~~~
okay cool, but this is for the entire summer experiment. when we make the stacked bar plots we want to group by disturbance, temperature, and week, so we need to do some more manipulations 
- The top_fam1$Percentage is summed across all samples, so it only tells  what fraction of the total dataset each Family contributes.For stacked bar plots,  each bar (e.g., each Treatment or each Temperature or each Week) should  sum to 100%. - Otherwise, bars won’t reflect relative community composition within that group.)

~~~~~~~~
created prep_bar_data() that:
- Groups data by a chosen variable (Treatment / Temperature / Week) and Family
- Sums abundances
- Converts them into relative abundance (%) within each group

Collapses rare structure using:
"Assigned at higher taxonomy" (families with underscores)
"Rare Families" (< 1%)
Otherwise keeps original family name

applied it to:
mdf_treatment
mdf_temperature
mdf_week
```{r}
library(dplyr)

library(dplyr)

# Function to prepare stacked bar plot data
prep_bar_data <- function(df, group_var) {
  
  df %>%
    group_by(.data[[group_var]], Family) %>%
    summarise(Abundance_sum = sum(Abundance), .groups = "drop") %>%
    
    group_by(.data[[group_var]]) %>%
    mutate(Percentage = Abundance_sum / sum(Abundance_sum) * 100) %>%
    ungroup() %>%
    
    mutate(Family_grouped = case_when(
      grepl("_", Family) ~ "Assigned at higher taxonomy",
      Percentage < 1 ~ "Rare Families",
      TRUE ~ Family
    ))
}

# Apply to each factor
mdf_treatment   <- prep_bar_data(mdffam.sum, "Treatment")
mdf_temperature <- prep_bar_data(mdffam.sum, "Temperature")
mdf_week        <- prep_bar_data(mdffam.sum, "Week")
```


assigning pallette colors! 
```{r}
palette_colors <- c(

  # Alphaproteobacteria (marine, surface-associated)
  "Rhodobacteraceae" = "#1f78b4",
  "Hyphomonadaceae"  = "#4ea3d9",

  # Gammaproteobacteria (heterotrophs, sulfur cycling, etc.)
  "Alteromonadaceae" = "#33a02c",
  "Marinobacteraceae"= "#66c2a5",
  "Halieaceae"       = "#41ae76",
  "Woeseiaceae"      = "#238b45",

  # Deltaproteobacteria (sulfate reducers)
  "Desulfobacteraceae"   = "#006d2c",
  "Desulfocapsaceae"     = "#31a354",
  "Desulfovibrionaceae"  = "#74c476",

  # Bacteroidetes (degraders, mats, organic matter)
  "Flavobacteriaceae" = "#984ea3",
  "Saprospiraceae"    = "#b07cc6",
  "Cryomorphaceae"    = "#c2a5cf",
  "Cyclobacteriaceae" = "#8c6bb1",
  "Bacteroidetes BD2-2" = "#d4b9da",

  # Planctomycetes / PVC group
  "Pirellulaceae"     = "#ff7f00",
  "Phycisphaeraceae"  = "#fdae6b",
  "Puniceicoccaceae"  = "#e6550d",

  # Cyanobacteria (phototrophs in mats)
  "Coleofasciculaceae" = "#e41a1c",
  "Nodosilineaceae"    = "#fb6a4a",

  # Firmicutes / other gram positives / thermophiles
  "Thermoanaerobaculaceae" = "#ffd92f",
  "Trueperaceae"           = "#e6ab02",

  # Spirochaetes
  "Spirochaetaceae" = "#a65628",

  # Misc /  environmental
  "A4b"                         = "#6a3d9a",
  "Acanthopleuribacteraceae"    = "#b15928",
  "Sandaracinaceae"             = "#b2df8a",
  "Solimonadaceae"              = "#cab2d6",
  "Micavibrionaceae"            = "#FF69B4",  # 💖 your pink (kept distinct on purpose)

  # Special 
  "Rare Families"                = "grey70",
  "Assigned at higher taxonomy"  = "darkblue",
  "Unknown Family"               = "black"
)
```

ensuring this worked, should say zero 
```{r}
setdiff(all_families, names(palette_colors))
```


```{r}
#Ordering treatment x axis
mdf_treatment$Treatment <- factor(
  mdf_treatment$Treatment,
  levels = c("ND", "Few Small", "Many Small", "Few Large", "HD")
)
```

```{r}
#  Alphabetical order (forced all plots to have the same stacking order/ color pallette)
family_levels <- sort(names(palette_colors))

# Apply to all datasets
mdf_treatment$Family_grouped   <- factor(mdf_treatment$Family_grouped, levels = family_levels)
mdf_temperature$Family_grouped <- factor(mdf_temperature$Family_grouped, levels = family_levels)
mdf_week$Family_grouped        <- factor(mdf_week$Family_grouped, levels = family_levels)
```


```{r}
# Plotting function ~ making the stakced bar plits using relative abundance
plot_bar <- function(df, xvar) {
  ggplot(df, aes(x = .data[[xvar]], y = Percentage, fill = Family_grouped)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(
      values = palette_colors,
      drop = FALSE
    ) +
    labs(
      x = xvar,
      y = "Relative Abundance (%)",
      fill = "Family"
    ) +
    theme_minimal(base_size = 16) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.text = element_text(size = 9),
      legend.title = element_text(size = 11)
    )
}
```


```{r}
#actual plotting
plot_treatment   <- plot_bar(mdf_treatment, "Treatment")
plot_temperature <- plot_bar(mdf_temperature, "Temperature")
plot_week        <- plot_bar(mdf_week, "Week")


# Extract ONE legend (treatment has all the taxa - 1, will add in the last one manually)

legend <- get_legend(
  plot_treatment +
    theme(
      legend.position = "bottom",
      legend.text = element_text(size = 8),
      legend.key.size = unit(0.3, "cm"),
      legend.spacing.y = unit(0.05, "cm"),
      legend.spacing.x = unit(0.05, "cm")
    ) +
    guides(fill = guide_legend(nrow = 5))
)

#  Remove legends from plots
plot_treatment_noleg   <- plot_treatment   + theme(legend.position = "none")
plot_temperature_noleg <- plot_temperature + theme(legend.position = "none")
plot_week_noleg        <- plot_week        + theme(legend.position = "none")


# cowplot combo
top_row <- plot_grid(
  plot_treatment_noleg,
  plot_temperature_noleg,
  plot_week_noleg,
  ncol = 3,
  rel_widths = c(.45, .4, .4)
)

f1bot <- plot_grid(
  top_row,
  legend,
  ncol = 1,
  rel_heights = c(.8,.2)
)

f1bot
```

```{r}
#double check, should say zero 
setdiff(sort(unique(c(
  mdf_treatment$Family_grouped,
  mdf_temperature$Family_grouped,
  mdf_week$Family_grouped
))), names(palette_colors))
```
```{r}
ggsave(
  filename = "fig1bot.pdf",
  plot = f1bot,
  width = 14,
  height = 10
)
```

DIFF ABUNDANCE ! ~ structural zero 
~results take 1-2 hours to run for each one~
upload raw phyloseq object 
```{r}
ps.summer.raw <-readRDS("ps.summer copy.raw.rds")
```

looking between treatments
```{r}
sum.diff.abund.treat <- ancombc2(
  data = ps.summer.raw,
  tax_level = "Family",
  fix_formula = "Treatment",
  p_adj_method = "holm",
  group = "Treatment",    # factor you are testing
  struc_zero = TRUE,        # detects structural zeros (presence/absence)
  neg_lb = TRUE,
  em_control = list(
    tol = 1e-5,
    max_iter = 100,
    sensitivity = TRUE      # runs subsampling stability check
  ),
  alpha = 0.05,
  global = TRUE             # adds global tests in addition to pairwise
)
```

```{r}
#structural zero results (need to clean up in excel)
View(sum.diff.abund.treat$zero_ind)
#write.csv(sum.diff.abund.treat$zero_ind, "sum_all_treat_strzeros.csv", row.names = TRUE)
#sensitivity results (scores are low, this is why we negate global and pairwise results and focus on the robust structural zero results above)
View(sum.diff.abund.treat$pseudo_sens_tab)
#write.csv(sum.diff.abund.treat$pseudo_sens_tab, "sum_all_treat_sens.csv", row.names = TRUE)
```

looking between temperatures
```{r}
sum.diff.abund.temp <- ancombc2(
  data = ps.summer.raw,
  tax_level = "Family",
  fix_formula = "Temperature",
  p_adj_method = "holm",
  group = "Temperature",    # factor you are testing
  struc_zero = TRUE,        # detects structural zeros (presence/absence)
  neg_lb = TRUE,
  em_control = list(
    tol = 1e-5,
    max_iter = 100,
    sensitivity = TRUE      # runs subsampling stability check
  ),
  alpha = 0.05,
  global = TRUE             # adds global tests in addition to pairwise
)
```

```{r}
#structural zero results (need to clean up in excel)
View(sum.diff.abund.temp$zero_ind)
#write.csv(sum.diff.abund.temp$zero_ind, "sum_all_temp_strzeros.csv", row.names = TRUE)
#sensitivity results (scores are low, this is why we negate global and pairwise results and focus on the robust structural zero results above)
View(sum.diff.abund.temp$pseudo_sens_tab)
#write.csv(sum.diff.abund.temp$pseudo_sens_tab, "sum_all_temp_sens.csv", row.names = TRUE)
```

```{r}
sum.diff.abund.weeks <- ancombc2(
  data = ps.summer.raw,
  tax_level = "Family",
  fix_formula = "Week",
  p_adj_method = "holm",
  group = "Week",    # factor you are testing
  struc_zero = TRUE,        # detects structural zeros (presence/absence)
  neg_lb = TRUE,
  em_control = list(
    tol = 1e-5,
    max_iter = 100,
    sensitivity = TRUE      # runs subsampling stability check
  ),
  alpha = 0.05,
  global = TRUE             # adds global tests in addition to pairwise
)
```

```{r}
#structural zero results (need to clean up in excel)
View(sum.diff.abund.weeks$zero_ind)
#write.csv(sum.diff.abund.weeks$zero_ind, "sum_all_weeks_strzeros.csv", row.names = TRUE)
#sensitivity results (scores are low, this is why we negate global and pairwise results and focus on the robust structural zero results above)
View(sum.diff.abund.weeks$pseudo_sens_tab)
#write.csv(sum.diff.abund.weeks$pseudo_sens_tab, "sum_all_weeks_sens.csv", row.names = TRUE)
```

final lab disturbance community composition (Figure 1) 
top: sum.treat.temp.week.ord
middle: sum.intplot1, sum.intplot2 (need to cowplot this)
bottom: f1bot

```{r}
library(cowplot)
sum.int.combo<- plot_grid( sum.intplot1,sum.intplot2, ncol = 2, labels = c("A", "B"))
sum.int.combo
```
okay now cowplot the whole thing! 
sum.treat.temp.week.ord, sum.int.combo, f1bot

```{r}
getwd()
```

```{r}
#FINAL figure 1. Will clean up in pdf/adobe
library(cowplot)

# Make top plot narrower
pA_narrow <- ggdraw() + 
  draw_plot(sum.treat.temp.week.ord, x = 0.5, y = 0.5, width = 0.7, hjust = 0.5, vjust = 0.5)

# Stack vertically
figure1 <- plot_grid(
  pA_narrow,        # narrow top plot
  sum.int.combo,     # middle plot (full width)
  f1bot,  # bottom plot (full width)
  ncol = 1,
  align = "v",
  rel_heights = c(1, 1.5, 2.5) # adjust vertical space as needed
)

figure1
```
```{r}
save_plot("figure1.pdf", figure1, 
          base_width =10 , base_height = 15)
```

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
Alpha Diversity Metrics
read in alpha diversity dataframes 

how did we make these? 
alpha_diversity<- data.frame(estimate_richness(ps.cleaner, split=TRUE, measures=c("Shannon","InvSimpson","Observed")))
alpha_diversity$Evenness <- alpha_diversity$Shannon / log(alpha_diversity$Observed)
^did this for each experiment and saved dataframes to be uploaded below 
ps.cleaner phyloseq object should not have transformed data (should use raw or minimally processed data for this)

```{r}
alpha.sum <- read.csv("alpha.div.sum.ch4.csv")
alpha.field <-read.csv("alpha.div.field.ch4.csv")
alpha.temp <- read.csv("alpha.div.temp.ch4.csv")
```
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
Lab dist (summer) alpha diversity stats

ensuring factoral variables
```{r}
alpha.sum$Treatment   <- as.factor(alpha.sum$Treatment)
alpha.sum$Temperature <- as.factor(alpha.sum$Temperature)
alpha.sum$Week        <- as.factor(alpha.sum$Week)
```

Shannon
```{r}
#maximal model (anova)
sumanov1 <- aov(Shannon ~ Treatment*Temperature*Week, data = alpha.sum)
summary(sumanov1)

#Assumption checks
#1. Normality
qqnorm(residuals(sumanov1))
qqline(residuals(sumanov1), col = "red")
shapiro_test <- shapiro.test(residuals(sumanov1))
print(shapiro_test)
#2. Homogeneity
library(car)
levene_test <- leveneTest(Shannon ~ Treatment*Temperature*Week, data = alpha.sum)
print(levene_test)
#3 linearity
plot(sumanov1, which = 1)
plot
#4 independence (think about sample design)
durbinWatsonTest(sumanov1)
```
- passes assumptions
- stat sig between weeks 

simple main effect & post hoc (week)
```{r}
sumanov1.2 <- aov(Shannon ~ Week, data = alpha.sum)
summary(sumanov1.2)
#posthoc test ~~ which weeks? 
TukeyHSD(sumanov1.2, which = "Week")
```
```{r}
#week plot for shannon div. in summer experiment
library(ggpubr)
library(ggplot2)

# Create plot with custom colors
shannon.sum <- ggplot(alpha.sum, aes(x = Week, y = Shannon, fill = Week)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "black") +
  scale_fill_manual(
    values = c(
      "1" = "tan4",
      "3" = "orange",
      "5" = "brown"
    )
  ) +
  coord_cartesian(ylim = c(5, 7.5)) +
  ylab("Shannon Diversity") +
  xlab("Week") +
  theme_classic(base_size = 14) +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 15),
    legend.position = "none"
  ) +
  stat_compare_means(
    comparisons = list(c("1", "5"), c("3", "5")), 
    method = "t.test", 
    label = "p.signif", 
    label.y = c(7.3, 7.5)  # adjust y positions as needed
  )

shannon.sum

# Save the plot
# ggsave("shannon.summer.alpha.png", plot = shannon.sum, width = 6, height = 5, dpi = 300)

```

Richness
```{r}
#maximal model
sumanov4 <- aov(Observed ~ Treatment*Temperature*Week, data = alpha.sum)
summary(sumanov4)

#Assumption checks
#1. Normality
qqnorm(residuals(sumanov4))
qqline(residuals(sumanov4), col = "red")
shapiro_test <- shapiro.test(residuals(sumanov4))
print(shapiro_test)
#2. Homogeneity
library(car)
levene_test <- leveneTest(Observed ~ Treatment*Temperature*Week, data = alpha.sum)
print(levene_test)
#3 linearity
plot(sumanov4, which = 1)
plot
#4 independence (think about sample design)
durbinWatsonTest(sumanov4)
```
untangle week:temp interaction here
- since alpha diversity is univariate data (and not multivariate distances like beta diversity), when we see an interaction, we can use emmeans *requires a single numeric response* 
- emmeans estimated marginal means, take fitted model (anova) and computes the mean response for each group, accounting for all other variables in the model 
```{r}
sumanovsimp <- aov(Observed ~  Temperature * Week, data = alpha.sum)
library(emmeans)
#estimated marginal means for the interaction
em1 <- emmeans(sumanovsimp, ~ Week | Temperature)
em2 <- emmeans(sumanovsimp, ~ Temperature | Week)
pairs(em1) 
pairs(em2) 
```

Interaction Plot:
```{r}
alpha.sum$Temperature <- factor(alpha.sum$Temperature, levels = c("22", "26"))
interactionplot <- ggplot(alpha.sum, aes(x = Week, y = Observed, color = Temperature, group = Temperature)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  scale_color_manual(values = c("22" = "blue3", "26" = "red3")) +
  theme_classic()
interactionplot
```
Inv Simpson *same anova pipeline above except with inv simpson, no stat differences 
Evenness **same anova pipeline above except with inv simpson, no stat differences

Figure 2 - Lab dist alpha div
```{r}
library(cowplot)
alpha.sum.com<- plot_grid( shannon.sum,interactionplot, 
                           labels = c("A", "B"),  # optional subplot labels
                           ncol = 1,               # number of columns
                           align = "v")           # align horizontally & vertically

# Display it
alpha.sum.com

```

~~~~~~~~~~~~~~~~~~
PAM 
```{r}
pam <- read.csv("PAM3.csv")
```

```{r}
#filter by experiment
sumpam <- pam[pam$Experiment == "Summer", ]
fieldpam <- pam[pam$Experiment == "Field", ]
temppam <- pam[pam$Experiment == "Temp", ]
```
~~~~~~~~~~~~~~~~~~~~
Summer PAM
-passed assumptions, no differences in pam between temperature or disturbance treatment
```{r}
#maximal model
sumpam1 <- aov(pam ~ Treatment * Temperature, data = sumpam)
summary(sumpam1)

#checkassumptions
#(1)normality
qqnorm(residuals(sumpam1))
qqline(residuals(sumpam1), col = "red")
shapiro_test <- shapiro.test(residuals(sumpam1))
print(shapiro_test)
#(2)homogeneity
sumpam$Week <- as.character(sumpam$Week)
sumpam$Temperature <- as.character(sumpam$Temperature)
levene_test <- leveneTest(pam ~ Treatment * Temperature, data = sumpam)
print(levene_test)
# (3) linearity~ Visual check: residuals vs fitted
plot(fitted(sumpam1), residuals(sumpam1),
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red")
# Look for random scatter around zero

```


~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
FIELD RESULTS! 

~Beta Dispersion
```{r}
field.bet.all <- betadisper(field.dist.all,field.samdf$Treatment)
field.bet.all
anova(field.bet.all)
```
*no stat difference in dispersion, what about beta diversity (differences in communities seen between treatment in situ)?

~Beta Diversity
```{r}
field.samdf$Treatment <- as.factor(field.samdf$Treatment)
adonis2(field.dist.all ~ Treatment , data=field.samdf, permutations=9999)
pairwise.adonis2(field.dist.all ~ Treatment , data = field.samdf, permutations = 9999)
```
No differences seen between any of the treatments in the field experiment

~Bar Plot
calculates the total abundance of each Family across all samples, sorts them by the rel abundance, and converts them into percentages of the total 

groups rare families into one group (<1% rel abundance) 
groups taxa with "_" which means it was assigned at a higher taxonomy 
```{r}
# Aggregate total relative abundance per Family across all samples
top_fam2 <- aggregate(Abundance ~ Family, data = mdffam.field, FUN = sum)

# Sort in descending order
top_fam2 <- top_fam2[order(-top_fam2$Abundance), ]

# Calculate the total abundance
total_abundance <- sum(top_fam2$Abundance)

# Calculate percentage for each Family
top_fam2$Percentage <- (top_fam2$Abundance / total_abundance) * 100

# Group rare families (<1%) and those lacking family-level assignment
top_fam2$Family_grouped <- ifelse(
  grepl("_", top_fam2$Family),             # condition 1: contains "_"
  "Assigned at higher taxonomy",           # assign this label
  ifelse(
    top_fam2$Percentage < 1,              # condition 2: <1% abundance
    "Rare Families",                       # assign this label
    top_fam2$Family                        # else keep original Family
  )
)

# View the result
print(top_fam2)
```
This is across all samples, now need to group by treatment and make a stacked bar plot
created prep_bar_data() that:
- Groups data by a chosen variable (Treatment) and Family
- Sums abundances
- Converts them into relative abundance (%) within each group

Collapses rare structure using:
"Assigned at higher taxonomy" (families with underscores)
"Rare Families" (< 1%)
Otherwise keeps original family name

applied it to:
mdf_treatment
```{r}
library(dplyr)

# Function to prepare stacked bar plot data
prep_bar_data2 <- function(df, group_var) {
  
  df %>%
    group_by(.data[[group_var]], Family) %>%
    summarise(Abundance_sum = sum(Abundance), .groups = "drop") %>%
    
    group_by(.data[[group_var]]) %>%
    mutate(Percentage = Abundance_sum / sum(Abundance_sum) * 100) %>%
    ungroup() %>%
    
    mutate(Family_grouped = case_when(
      grepl("_", Family) ~ "Assigned at higher taxonomy",
      Percentage < 1 ~ "Rare Families",
      TRUE ~ Family
    ))
}

# Apply to each factor
mdf_treatment2   <- prep_bar_data2(mdffam.field, "Treatment")
```

making an "all families" vector
```{r}
all_families2 <- unique(c(
  mdf_treatment2$Family_grouped))
```

looking at the different families to make a color pallette 
```{r}
unique(mdf_treatment2$Family_grouped)
```
creating a similar color pallette as used in the summer experiment: 
```{r}
palette_colors2 <- c(

  # Alphaproteobacteria
  "Rhodobacteraceae" = "#1f78b4",

  # Gammaproteobacteria (incl. sulfur oxidizers)
  "Halieaceae"       = "#41ae76",
  "Woeseiaceae"      = "#238b45",
  "Thiotrichaceae"   = "#74c476",  # sulfur oxidizers → lighter green
  "Chromatiaceae"    = "#006d2c",  # purple sulfur bacteria → deep green

  # Deltaproteobacteria (sulfate reducers)
  "Desulfobacteraceae"  = "#00441b",
  "Desulfocapsaceae"    = "#31a354",
  "Desulfovibrionaceae" = "#74c476",
  "Desulfobulbaceae"    = "#238b45",
  "Desulfosarcinaceae"  = "#006d2c",

  # Bacteroidetes
  "Flavobacteriaceae"   = "#984ea3",
  "Saprospiraceae"      = "#b07cc6",
  "Cyclobacteriaceae"   = "#8c6bb1",
  "Bacteroidetes BD2-2" = "#d4b9da",
  "Prolixibacteraceae"  = "#c2a5cf",

  # Planctomycetes / PVC
  "Pirellulaceae"     = "#ff7f00",
  "Puniceicoccaceae"  = "#e6550d",

  # Firmicutes / thermophiles / anaerobes
  "Thermoanaerobaculaceae" = "#ffd92f",

  # Spirochaetes
  "Spirochaetaceae" = "#a65628",

  # Misc / environmental
  "A4b" = "#6a3d9a",

  # Special
  "Rare Families"               = "grey70",
  "Assigned at higher taxonomy" = "darkblue",
  "Unknown Family"              = "black"
)
```

ensuring this worked, should say zero 
```{r}
setdiff(all_families2, names(palette_colors2))
```


```{r}
#Ordering treatment x axis
mdf_treatment2$Treatment <- factor(
  mdf_treatment2$Treatment,
  levels = c("ND", "Few Small", "Many Small", "HD")
)
```

```{r}
#  Alphabetical order (forced all plots to have the same stacking order/ color pallette)
family_levels2 <- sort(names(palette_colors2))

# Apply to all datasets
mdf_treatment2$Family_grouped   <- factor(mdf_treatment2$Family_grouped, levels = family_levels2)
```


```{r}
# Plotting function ~ making the stacked bar plots using relative abundance
plot_bar <- function(df, xvar) {
  ggplot(df, aes(x = .data[[xvar]], y = Percentage, fill = Family_grouped)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(
      values = palette_colors2,
      drop = FALSE
    ) +
    labs(
      x = xvar,
      y = "Relative Abundance (%)",
      fill = "Family"
    ) +
    theme_minimal(base_size = 16) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.text = element_text(size = 9),
      legend.title = element_text(size = 11)
    )
}
```


```{r}
#actual plotting
plot_treatment2   <- plot_bar(mdf_treatment2, "Treatment")
plot_treatment2
```
field.treat.ord
plot_treatment2

combo field plot
```{r}
#FINAL COMBO FIELD PLOT
library(cowplot)

# Stack the two plots vertically with labels
combo_field <- plot_grid(
  field.treat.ord,
  plot_treatment2,
  nrow = 2,          # vertical layout
  align = "v",       # align plots vertically
  label_fontface = "bold",   # bold labels
  label_x = 0.02,            # left-align labels
  label_y = 0.98             # position near top
)

combo_field
```

```{r}
save_plot("finalfieldplot.pdf", combo_field, 
          base_width =8 , base_height = 8)
```

~Diff Abundance Results

```{r}
field.diff.abund <- ancombc2(
  data = ps.field.raw,
  tax_level = "Family",
  fix_formula = "Treatment",
  p_adj_method = "holm",
  group = "Treatment",    # factor you are testing
  struc_zero = TRUE,        # detects structural zeros (presence/absence)
  neg_lb = TRUE,
  em_control = list(
    tol = 1e-5,
    max_iter = 100,
    sensitivity = TRUE      # runs subsampling stability check
  ),
  alpha = 0.05,
  global = TRUE             # adds global tests in addition to pairwise
)
```

```{r}
#structural zero results (need to clean up in excel)
View(field.diff.abund$zero_ind)
#write.csv(field.diff.abund$zero_ind, "field_strzeros.csv", row.names = TRUE)
#sensitivity results (scores are low, this is why we negate global and pairwise results and focus on the robust structural zero results above)
View(field.diff.abund$pseudo_sens_tab)
#write.csv(field.diff.abund$pseudo_sens_tab, "field_sens.csv", row.names = TRUE)
```

~Alpha Diversity Models
alpha.field
Shannon, inv simpson, evenness, richness

*did this for all alpha div metrics (above), all models passed assumptions, no stat sig results seen 
```{r}
alpha.field$Treatment   <- as.character(alpha.field$Treatment)

#maximal model
fieldanov1 <- aov(Shannon ~ Treatment, data = alpha.field)
summary(fieldanov1)

#Assumption checks
#1. Normality
qqnorm(residuals(fieldanov1))
qqline(residuals(fieldanov1), col = "red")
shapiro_test <- shapiro.test(residuals(fieldanov1))
print(shapiro_test)
#2. Homogeneity
library(car)
levene_test <- leveneTest(Shannon ~ Treatment, data = alpha.field)
print(levene_test)
#3 linearity
plot(fieldanov1, which = 1)
plot
#4 independence (think about sample design)
durbinWatsonTest(fieldanov1)
```

~PAM
Field PAM, just a baseline, looked at internal vs external samples (no between treatments)
~no differences seen
```{r}

fieldpam$Treatment <- as.character(fieldpam$Treatment)
fieldpam_model <- aov(pam ~ Treatment, data = fieldpam)
summary(fieldpam_model)

#checkassumptions
#(1)normality
qqnorm(residuals(fieldpam_model))
qqline(residuals(fieldpam_model), col = "red")
shapiro_test <- shapiro.test(residuals(fieldpam_model))
print(shapiro_test)
#(2)homogeneity
levene_test <- leveneTest(pam ~ Treatment, data = fieldpam)
print(levene_test)
```
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~Temperature Experiment

~Beta dispersion
```{r}
#temp
temp.bet.all1 <- betadisper(temp.dist.all,temp.samdf$Temperature)
anova(temp.bet.all1)
```
*no differences in dispersion between temperatures


```{r}
#week
temp.bet.all2 <- betadisper(temp.dist.all,temp.samdf$Week)
anova(temp.bet.all2)
TukeyHSD(temp.bet.all2)
```
*differences between weeks (3 vs 1 and 5 vs 1) ~ keep this in mind for beta diveristy results

~ Beta Diveristy

full model! 
  *including weeks
```{r}
temp.samdf$Temperature <- as.factor(temp.samdf$Temperature)
temp.samdf$Week <- as.factor(temp.samdf$Week)
#temp.dist.all <- vegdist(temp.seqtab, method = "euclidean")
adonis2(temp.dist.all ~ Temperature*Week , data=temp.samdf, permutations=9999)
```
~Tease apart interaction

in which weeks to temps differ? 
```{r}
# Get unique week levels
weeks <- unique(temp.samdf$Week)

# Loop over each week to test for temperature effect
for (w in weeks) {
  
  cat("\n--- PERMANOVA: Temperature differences within Week =", w, "---\n")
  
  # Logical index for samples matching the current week
  week_idx <- temp.samdf$Week == w
  
  # Subset distance matrix and metadata
  dist_subset <- as.dist(as.matrix(temp.dist.all)[week_idx, week_idx])
  meta_subset <- temp.samdf[week_idx, ]
  
  # Check if at least 2 temperature levels are present
  if (length(unique(meta_subset$Temperature)) > 1) {
    # Run PERMANOVA
    result <- adonis2(dist_subset ~ Temperature, data = meta_subset, permutations = 9999)
    print(result)
  } else {
    cat("Not enough temperature levels to compare for this week.\n")
  }
}

```
Temperatures differ in weeks 1, 3, and 5!

at which temps do week differ? 
```{r}
# Get unique temperature levels
temps <- unique(temp.samdf$Temperature)

# Loop over each temperature to test for week effect
for (temp in temps) {
  
  cat("\n--- PERMANOVA: Week differences within Temperature =", temp, "---\n")
  
  # Logical index for samples matching the current temperature
  temp_idx <- temp.samdf$Temperature == temp
  
  # Subset distance matrix and metadata
  dist_subset <- as.dist(as.matrix(temp.dist.all)[temp_idx, temp_idx])
  meta_subset <- temp.samdf[temp_idx, ]
  
  # Check if at least 2 week levels are present
  if (length(unique(meta_subset$Week)) > 1) {
    # Run PERMANOVA
    result <- adonis2(dist_subset ~ Week, data = meta_subset, permutations = 9999)
    print(result)
  } else {
    cat("Not enough week levels to compare for this temperature.\n")
  }
}
```
Yup, weeks differ at all temperature treatments as well
*pairwise tests cannot be completed, sample size too small, so let's look at trends with interaction plots! 

~interaction plots 
```{r}
ordination_df <- read.csv("ordination_df.csv")
```

```{r}
#FINAL interaction plot 1 for temperature 
p.2 = ggplot(ordination_df, aes(x = as.factor(Week), y = PCoA1, color = as.factor(Temperature), group = Temperature)) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  scale_color_manual(
    values = c(
      "26" = "blue3",
      "29" = "darkgreen",
      "31" = "magenta4",
      "33" = "red3"
    )
  ) +
  labs(
    title = "",
    x = "Week",
    y = "PCoA Axis 1 score",
    color = "Temperature"
  ) +
  theme_minimal()+
  theme(
    axis.title = element_text(size = 16),  # axis titles
    axis.text = element_text(size = 14),                  # axis tick labels
    legend.text = element_text(size = 14),                # legend labels
    legend.title = element_text(size = 16)                # legend title
  )
p.2
#ggsave("interaction.beta,div.pdf", plot = p.2, width = 8, height = 6, units = "in", dpi = 300)
```

```{r}
#FINAL interaction plot 2 for temperature 

p.2.5 = ggplot(ordination_df, aes(x = as.factor(Temperature), y = PCoA1, color = as.factor(Week), group = Week)) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  scale_color_manual(
    values = c(
      "1" = "tan4",
      "3" = "orange",
      "5" = "brown"
    )
  ) +
  labs(
    title = "",
    x = "Temperature",
    y = "PCoA Axis 1 score",
    color = "Week"
  ) +
 theme_minimal()+
  theme(
    axis.title = element_text(size = 16),  # axis titles
    axis.text = element_text(size = 14),                  # axis tick labels
    legend.text = element_text(size = 14),                # legend labels
    legend.title = element_text(size = 16)                # legend title
  )

p.2.5

```
~ Reduced temperature model & pairwise results (supports interaction plot results)
```{r}
adonis2(temp.dist.all ~ Temperature , data = temp.samdf, permutations = 9999)
pairwise.adonis2(temp.dist.all ~ Temperature , data = temp.samdf, permutations = 9999)
```
~Bar plots 

calculates the total abundance of each Family across all samples, sorts them by the rel abundance, and converts them into percentages of the total 

groups rare families into one group (<1% rel abundance) 
groups taxa with "_" which means it was assigned at a higher taxonomy 
```{r}
# Aggregate total relative abundance per Family across all samples
top_fam2 <- aggregate(Abundance ~ Family, data = mdffam.field, FUN = sum)

# Sort in descending order
top_fam2 <- top_fam2[order(-top_fam2$Abundance), ]

# Calculate the total abundance
total_abundance <- sum(top_fam2$Abundance)

# Calculate percentage for each Family
top_fam2$Percentage <- (top_fam2$Abundance / total_abundance) * 100

# Group rare families (<1%) and those lacking family-level assignment
top_fam2$Family_grouped <- ifelse(
  grepl("_", top_fam2$Family),             # condition 1: contains "_"
  "Assigned at higher taxonomy",           # assign this label
  ifelse(
    top_fam2$Percentage < 1,              # condition 2: <1% abundance
    "Rare Families",                       # assign this label
    top_fam2$Family                        # else keep original Family
  )
)

# View the result
print(top_fam2)
```
