---
title: "DisturbanceChapter_GitHub"
output: html_document
date: "2026-04-06"
---
Summarized Code/Clean Code: 

Community Composition:

loading libraries
```{r}
library(stringr)
library(dplyr)
library(stats)
library(ggplot2)
library(ggpubr)
library(vegan)
library(cowplot)
library(tidyverse)
library(phyloseq)
library(microbiome)
library("funfuns")
library(pairwiseAdonis)
library(viridis)
library("microViz")
```

```{r}
getwd()
```

read in clean phyloseq object & seqtab object from dada2pipeline with rel abundance counts applied
(ps.cleaner.rel)

```{r}
# load it back into R
ps.cleaner.clr <- readRDS("ch4.ps.cleaner.clr.rds")
```

applying a CLR (Centered Log-Ratio) transformation 
making a sequence table (already loaded in clr version above, so skip this for now)
```{r}
ps.cleaner.clr <- microbiome::transform(ps.cleaner.rel,'clr')
seqtab.clr <- data.frame(otu_table(ps.cleaner.clr))
```

so....
- ps.cleaner.rel --> relative abundances of sample counts
- ps.cleaner.clr --> relative abundances with the clr transformation 
- seqtab.clr --> clr transformed sequence table

subsetting into 3 experiments: 
```{r}
ps.summer <- subset_samples(ps.cleaner.clr, Experiment == "Summer")
ps.field <- subset_samples(ps.cleaner.clr, Experiment == "Field")
ps.temp <- subset_samples(ps.cleaner.clr, Experiment == "Temp")
```

PCOA plots to visualize spread before stats!
~Lab Dist (Summer)
figure 1a
```{r}
# Make Treatment a factor in the desired order
ps.summer@sam_data$Treatment <- factor(
  ps.summer@sam_data$Treatment,
  levels = c("ND", "Few Small", "Many Small", "Few Large", "HD")
)

# Now plot
sum.treat.temp.week.ord <- plot_ordination(
  ps.summer,
  ordinate(ps.summer, "PCoA", "euclidean")
) +
  geom_point(
    aes(
      fill = Treatment,
      shape = as.factor(Temperature)
    ),
    size = 4,
    color = "black",
    alpha = 0.8
  ) +
  stat_ellipse(
    aes(group = as.factor(Week), color = as.factor(Week)),
    linetype = "dashed",
    size = 1
  ) +
  scale_color_manual(
    name = "Week",
    values = c("1" = "tan4", "3" = "orange", "5" = "brown")
  ) +
  scale_fill_manual(
    name = "Treatment",
    values = c(
      "ND" = "#999999",
      "Few Small" = "#d95f02",
      "Many Small" = "#0072B2",
      "Few Large" = "#1b9e77",
      "HD" = "purple3"
    )
  ) +
  scale_shape_manual(
    name = "Temperature (°C)",
    values = c("22" = 21, "26" = 22, "30" = 24)
  ) +
  guides(
    fill = guide_legend(
      override.aes = list(shape = 21),
      keywidth = 0.4, keyheight = 0.4
    ),
    shape = guide_legend(
      override.aes = list(fill = "white"),
      keywidth = 0.4, keyheight = 0.4
    )
  ) +
  theme_classic(base_size = 22) +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.spacing.y = unit(0.0005, "cm"),
    legend.key.size = unit(0.4, "lines")
  )

sum.treat.temp.week.ord
```

~Field Dist Experiment(Field)
figure 3a
```{r}
# Make Treatment a factor
sample_data(ps.field)$Treatment <- factor(
  sample_data(ps.field)$Treatment,
  levels = c("ND", "Few Small", "Many Small", "HD")
)

# Ordination
ord <- ordinate(ps.field, method = "PCoA", distance = "euclidean")

# Plot
field.treat.ord <- plot_ordination(ps.field, ord) +
  geom_point(
    aes(fill = Treatment),
    size = 4,
    shape = 21,
    color = "black",
    alpha = 0.8
  ) +
  stat_ellipse(
   aes(group = Treatment, color = Treatment),
  linetype = "dashed",
   size = 1
  ) +
  theme_classic(base_size = 22) +
  scale_fill_manual(
    name = "Treatment",
    values = c(
      "ND" = "#999999",
      "Few Small" = "#d95f02",
      "Many Small" = "#0072B2",
      "HD" = "purple3"
    )
  ) +
  
  scale_color_manual(
    name = "Treatment",
    values = c(
"ND" = "#999999",
      "Few Small" = "#d95f02",
      "Many Small" = "#0072B2",
      "HD" = "purple3")
  ) +
  
  guides(
    fill = guide_legend(override.aes = list(shape = 21)),
    color = "none"
  ) +
  theme(
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16)
  )

field.treat.ord
```

~ Temperature Experiment (Temp)
figure 4a
```{r}
# Make factors
sample_data(ps.temp)$Week <- factor(sample_data(ps.temp)$Week, levels = c("1", "3", "5"))
sample_data(ps.temp)$Temperature <- factor(sample_data(ps.temp)$Temperature, levels = c("26", "29", "31", "33"))

# Ordination
ord <- ordinate(ps.temp, method = "PCoA", distance = "euclidean")

# Plot
temp.temp.week.ord <- plot_ordination(ps.temp, ord) +
  geom_point(
    aes(color = Temperature, shape = Week),
    size = 4,
    alpha = 0.8
  ) +
  stat_ellipse(
    aes(group = Week, color = Week),  # Ellipse by Week
    linetype = "dashed",
    size = 1
  ) +
  scale_color_manual(
    values = c(
      "26" = "blue3", "29" = "darkgreen", "31" = "magenta4", "33" = "red3",
      "1" = "tan4", "3" = "orange", "5" = "brown"  # for Week ellipses
    )
  ) +
  guides(
    shape = guide_legend(override.aes = list(color = "black")), # Week shapes
    color = guide_legend(override.aes = list(shape = 16))        # Temp colors
  ) +
  theme_classic(base_size = 22) +
  theme(
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 18),
    legend.text = element_text(size = 18),
    legend.title = element_text(size = 18)
  )

temp.temp.week.ord
#good enough, will edit key later
```

Set up for Beta Dispersion

-subset seqtab by experiment (sequence table)
```{r}
sum.seqtab <- readRDS("chap4sumseqtab.clr.rds")
field.seqtab <- readRDS("chap4fieldseqtab.clr.rds")
temp.seqtab <- readRDS("chap4tempseqtab.clr.rds")
```

-also samdf is all the metadata, this needs to be read it and subset by experiment as well (already subsetted, so just reading in) (metadata)
```{r}
sum.samdf <- read.csv("chap4sum.samdf.csv")
field.samdf <- read.csv("chap4field.samdf.csv")
temp.samdf <- read.csv("chap4temp.samdf.csv")
```

Make a Distance Matrix: veg-distance calculates pairwise distances between samples (using euclidean distances)
```{r}
sum.dist.all <- vegdist(sum.seqtab, method = "euclidean")
field.dist.all <- vegdist(field.seqtab, method = "euclidean")
temp.dist.all <- vegdist(temp.seqtab, method = "euclidean")
```

Beta Dispersion: 
*looks at spread of points in a multidemensional space
*is beta dispersion differs a lot, PERMANOVA results may be biased

~Summer

```{r}
sum.bet.all1 <- betadisper(sum.dist.all,sum.samdf$Temperature)
sum.bet.all2 <- betadisper(sum.dist.all,sum.samdf$Week)
sum.bet.all3 <- betadisper(sum.dist.all,sum.samdf$Treatment)
```

```{r}
#temp
sum.bet.all1
anova(sum.bet.all1)
```
*no differences in dispersion between temps

```{r}
#week
sum.bet.all2
anova(sum.bet.all2)
TukeyHSD(sum.bet.all2)
```
*diff between:
- weeks 5 & 3
- weeks 5 & 1 


```{r}
#treatment
sum.bet.all3
anova(sum.bet.all3)
```
*no differences in dispersion between treatments

~Field
```{r}
field.bet.all <- betadisper(field.dist.all,field.samdf$Treatment)
field.bet.all
anova(field.bet.all)
```
*no differences in dispersion between  treatments

~Temp
```{r}
#temp
temp.bet.all1 <- betadisper(temp.dist.all,temp.samdf$Temperature)
anova(temp.bet.all1)
```
*no differences in dispersion betweentemperatures


```{r}
#week
temp.bet.all2 <- betadisper(temp.dist.all,temp.samdf$Week)
anova(temp.bet.all2)
TukeyHSD(temp.bet.all2)
```
*diff between:
- weeks 3 & 1
- weeks 5 & 1 


Beta Diversity (PERMANOVAS)

~ summer full model 
```{r}
adonis2(sum.dist.all ~ Treatment * Temperature *Week , data=sum.samdf, permutations=9999)
```
Treatment                   4    90414 0.08753 1.7698 0.0001 ***
Temperature                 1    23652 0.02290 1.8519 0.0021 ** 
Week                        1    59115 0.05723 4.6286 0.0001 ***
Treatment:Temperature       3    53248 0.05155 1.3897 0.0026 ** 

Since treatment:temp interaction, need to tear apart interaction 
week has a significant difference, may be influence by beta dispersion 

subset data ~ are treatments different at 22C only? YES
Example:
```{r}
# Logical index for samples at Temperature 22
temp22_idx <- sum.samdf$Temperature == "22"

# Subset the distance matrix (use square brackets once)
dist_temp22 <- as.dist(as.matrix(sum.dist.all)[temp22_idx, temp22_idx])

# Subset metadata to match
meta_temp22 <- sum.samdf[temp22_idx, ]

# Run PERMANOVA
adonis2(dist_temp22 ~ Treatment, data = meta_temp22, permutations = 9999)


```
subset data ~ are treatments different at 26C only? YES
```{r}
# Logical index for samples at Temperature 22
temp26_idx <- sum.samdf$Temperature == "26"

# Subset the distance matrix (use square brackets once)
dist_temp26 <- as.dist(as.matrix(sum.dist.all)[temp26_idx, temp26_idx])

# Subset metadata to match
meta_temp26 <- sum.samdf[temp26_idx, ]

# Run PERMANOVA
adonis2(dist_temp26 ~ Treatment, data = meta_temp26, permutations = 9999)
```
subset data ~ are temperatures different between the treatments? 
Yes, for ND & Few Small 
No, for Many Small & Few Large 
```{r}
# Get unique treatment levels
treatments <- unique(sum.samdf$Treatment)

# Loop over each treatment to test for temperature effect
for (treat in treatments) {
  
  cat("\n--- PERMANOVA: Temperature differences within Treatment =", treat, "---\n")
  
  # Logical index for samples matching the treatment
  treat_idx <- sum.samdf$Treatment == treat
  
  # Subset distance matrix and metadata
  dist_subset <- as.dist(as.matrix(sum.dist.all)[treat_idx, treat_idx])
  meta_subset <- sum.samdf[treat_idx, ]
  
  # Check if at least 2 temperature levels are present
  if (length(unique(meta_subset$Temperature)) > 1) {
    # Run PERMANOVA
    result <- adonis2(dist_subset ~ Temperature, data = meta_subset, permutations = 9999)
    print(result)
  } else {
    cat("Not enough temperature levels to compare for this treatment.\n")
  }
}

```

interaction plots for summer (lab) experiment


```{r}
ordination_df2 <- read.csv("ordination_df2.csv")
```

figure 1B (left)
```{r}
#final interaction plot 1 for summer
ordination_df2$Treatment <- factor(
  ordination_df2$Treatment,
  levels = c("ND", "Few Small", "Many Small", "Few Large", "HD")
)
sum.intplot1 <- ggplot(ordination_df2, aes(
  x = as.factor(Temperature),
  y = PCoA1,
  color = Treatment,
  group = Treatment
)) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  scale_color_manual(values = c(
    "ND" = "#999999",        # neutral gray
    "Few Small" = "#d95f02", # orange
    "Many Small" = "#0072B2",# magenta-pink
    "Few Large" = "#1b9e77", # teal-green
    "HD" = "purple3"         # purple
  )) +
  labs(
    title = "",
    x = "Temperature",
    y = "PCoA Axis 1 score",
    color = "Treatment"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 14)
  )

sum.intplot1


```
figure 1B (right)
```{r}
# Then create the plot
sum.intplot2 <- ggplot(ordination_df2, aes(x = Treatment, y = PCoA1, color = as.factor(Temperature), group = Temperature)) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  scale_color_manual(
    values = c(
      "22" = "blue3",
      "26" = "red3"
    )
  ) +
  labs(
    title = "",
    x = "Grazing Treatment",
    y = "PCoA Axis 1 score",
    color = "Temperature"
  ) +
  
   theme_minimal(base_size = 16) +   # ← increases all font sizes (default = 11)
  theme(
    axis.title = element_text(size = 20),
    axis.text = element_text(size = 13),
    legend.title = element_text(size = 20),
    legend.text = element_text(size = 17)
  )


sum.intplot2
```



week alone (simple main effect)

```{r}
adonis2(sum.dist.all ~ Week , data=sum.samdf, permutations=9999)
pairwise.adonis2(sum.dist.all ~ Week , data = sum.samdf, permutations = 9999)
```

Bar plots *updated

upload mdf files, what are they?  a tidy data frame of relative abundances per sample and taxonomic group, formatted for plotting stacked bar plots ~ we want to group the stacked bar plots so we don't have a bunch of rare species, lets look
```{r}
mdffam.sum <- read.csv("ch4.mdffam.summer.csv")
mdffam.field <- read.csv("ch4.mdffam.field.csv")
mdffam.temp <- read.csv("ch4.mdffam.temp.csv")
```

calculates the total abundance of each Family across all samples, sorts them by the rel abundance, and converts them into percentages of the total 

groups rare families into one group (<1% rel abundance) 

groups taxa with "_" which means it was assigned at a higher taxonomy 
```{r}
# Aggregate total relative abundance per Family across all samples
top_fam1 <- aggregate(Abundance ~ Family, data = mdffam.sum, FUN = sum)

# Sort in descending order
top_fam1 <- top_fam1[order(-top_fam1$Abundance), ]

# Calculate the total abundance
total_abundance <- sum(top_fam1$Abundance)

# Calculate percentage for each Family
top_fam1$Percentage <- (top_fam1$Abundance / total_abundance) * 100

# Group rare families (<1%) and those lacking family-level assignment
top_fam1$Family_grouped <- ifelse(
  grepl("_", top_fam1$Family),             # condition 1: contains "_"
  "Assigned at higher taxonomy",           # assign this label
  ifelse(
    top_fam1$Percentage < 1,              # condition 2: <1% abundance
    "Rare Families",                       # assign this label
    top_fam1$Family                        # else keep original Family
  )
)

# View the result
print(top_fam1)
```

okay cool, but this is for the entire summer experiment. when we make the stacked bar plots we want to group by disturbance, temperature, and week, so we need to do some more manipulations 

- The top_fam1$Percentage is summed across all samples, so it only tells  what fraction of the total dataset each Family contributes.
- For stacked bar plots,  each bar (e.g., each Treatment or each Temperature or each Week) should  sum to 100%. Otherwise, bars won’t reflect relative community composition within that group.

```{r}
library(dplyr)

library(dplyr)

# Function to prepare stacked bar plot data
prep_bar_data <- function(df, group_var) {
  
  df %>%
    group_by(.data[[group_var]], Family) %>%
    summarise(Abundance_sum = sum(Abundance), .groups = "drop") %>%
    
    group_by(.data[[group_var]]) %>%
    mutate(Percentage = Abundance_sum / sum(Abundance_sum) * 100) %>%
    ungroup() %>%
    
    mutate(Family_grouped = case_when(
      grepl("_", Family) ~ "Assigned at higher taxonomy",
      Percentage < 1 ~ "Rare Families",
      TRUE ~ Family
    ))
}

# Apply to each factor
mdf_treatment   <- prep_bar_data(mdffam.sum, "Treatment")
mdf_temperature <- prep_bar_data(mdffam.sum, "Temperature")
mdf_week        <- prep_bar_data(mdffam.sum, "Week")
```

let's make 3 stacked bar plots with the same color scheme!
Figure 1C

```{r}
# Get all unique Family_grouped levels across all datasets
all_families <- unique(c(
  mdf_treatment$Family_grouped,
  mdf_temperature$Family_grouped,
  mdf_week$Family_grouped
))
```
```{r}
library(scales)
palette_colors <- hue_pal(h = c(0, 360) + 15, c = 100, l = 65)(length(all_families))
names(palette_colors) <- all_families

palette_colors["Rare Families"] <- "grey70"
palette_colors["Assigned at higher taxonomy"] <- "darkblue"
palette_colors["Unknown Family"] <- "darkblue"

```

```{r}
library(ggplot2)

plot_bar <- function(df, xvar) {
  ggplot(df, aes(x = .data[[xvar]], y = Percentage, fill = Family_grouped)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = palette_colors) +
    labs(
      x = xvar,
      y = "Relative Abundance (%)",
      fill = "Family"
    ) +
    theme_minimal(base_size = 16) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.text = element_text(size = 10),
      legend.title = element_text(size = 12)
    )
}

plot_treatment   <- plot_bar(mdf_treatment, "Treatment")
plot_temperature <- plot_bar(mdf_temperature, "Temperature")
plot_week        <- plot_bar(mdf_week, "Week")
```

```{r}
plot_treatment
```

```{r}
plot_temperature <- plot_bar(mdf_temperature, "Temperature")
plot_temperature
```
```{r}
plot_week<- plot_bar(mdf_week, "Week")
plot_week
```

