---
title: "DisturbanceChapter_GitHub"
output: html_document
date: "2026-04-06"
---
Summarized Code:

Community Composition:

loading libraries
```{r}
library(stringr)
library(dplyr)
library(stats)
library(ggplot2)
library(ggpubr)
library(vegan)
library(cowplot)
library(tidyverse)
library(phyloseq)
library(microbiome)
library("funfuns")
library(pairwiseAdonis)
library(viridis)
library("microViz")
```

read in clean phyloseq object & seqtab object from dada2pipeline with rel abundance counts applied
(ps.cleaner.rel)

applying a CLR (Centered Log-Ratio) transformation 
making a sequence table
```{r}
ps.cleaner.clr <- microbiome::transform(ps.cleaner.rel,'clr')
seqtab.clr <- data.frame(otu_table(ps.cleaner.clr))
```

so....
- ps.cleaner.rel --> relative abundances of sample counts
- ps.cleaner.clr --> relative abundances with the clr transformation 
- seqtab.clr --> clr transformed sequence table

subsetting into 3 experiments: 
```{r}
ps.summer <- subset_samples(ps.cleaner.clr, Experiment == "Summer")
ps.field <- subset_samples(ps.cleaner.clr, Experiment == "Field")
ps.temp <- subset_samples(ps.cleaner.clr, Experiment == "Temp")
```

PCOA plots to visualize spread before stats!
~Lab Dist (Summer)
```{r}
# Make Treatment a factor in the desired order
ps.summer@sam_data$Treatment <- factor(
  ps.summer@sam_data$Treatment,
  levels = c("ND", "Few Small", "Many Small", "Few Large", "HD")
)

# Now plot
sum.treat.temp.week.ord <- plot_ordination(
  ps.summer,
  ordinate(ps.summer, "PCoA", "euclidean")
) +
  geom_point(
    aes(
      fill = Treatment,
      shape = as.factor(Temperature)
    ),
    size = 4,
    color = "black",
    alpha = 0.8
  ) +
  stat_ellipse(
    aes(group = as.factor(Week), color = as.factor(Week)),
    linetype = "dashed",
    size = 1
  ) +
  scale_color_manual(
    name = "Week",
    values = c("1" = "tan4", "3" = "orange", "5" = "brown")
  ) +
  scale_fill_manual(
    name = "Treatment",
    values = c(
      "ND" = "#999999",
      "Few Small" = "#d95f02",
      "Many Small" = "#0072B2",
      "Few Large" = "#1b9e77",
      "HD" = "purple3"
    )
  ) +
  scale_shape_manual(
    name = "Temperature (°C)",
    values = c("22" = 21, "26" = 22, "30" = 24)
  ) +
  guides(
    fill = guide_legend(
      override.aes = list(shape = 21),
      keywidth = 0.4, keyheight = 0.4
    ),
    shape = guide_legend(
      override.aes = list(fill = "white"),
      keywidth = 0.4, keyheight = 0.4
    )
  ) +
  theme_classic(base_size = 22) +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.spacing.y = unit(0.0005, "cm"),
    legend.key.size = unit(0.4, "lines")
  )

sum.treat.temp.week.ord
```

~Field Dist Experiment(Field)
```{r}
# Make Treatment a factor
sample_data(ps.field)$Treatment <- factor(
  sample_data(ps.field)$Treatment,
  levels = c("ND", "Few Small", "Many Small", "HD")
)

# Ordination
ord <- ordinate(ps.field, method = "PCoA", distance = "euclidean")

# Plot
field.treat.ord <- plot_ordination(ps.field, ord) +
  geom_point(
    aes(fill = Treatment),
    size = 4,
    shape = 21,
    color = "black",
    alpha = 0.8
  ) +
  stat_ellipse(
   aes(group = Treatment, color = Treatment),
  linetype = "dashed",
   size = 1
  ) +
  theme_classic(base_size = 22) +
  scale_fill_manual(
    name = "Treatment",
    values = c(
      "ND" = "#999999",
      "Few Small" = "#d95f02",
      "Many Small" = "#0072B2",
      "HD" = "purple3"
    )
  ) +
  
  scale_color_manual(
    name = "Treatment",
    values = c(
"ND" = "#999999",
      "Few Small" = "#d95f02",
      "Many Small" = "#0072B2",
      "HD" = "purple3")
  ) +
  
  guides(
    fill = guide_legend(override.aes = list(shape = 21)),
    color = "none"
  ) +
  theme(
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16)
  )

field.treat.ord
```

~ Temperature Experiment (Temp)
```{r}
# Make factors
sample_data(ps.temp)$Week <- factor(sample_data(ps.temp)$Week, levels = c("1", "3", "5"))
sample_data(ps.temp)$Temperature <- factor(sample_data(ps.temp)$Temperature, levels = c("26", "29", "31", "33"))

# Ordination
ord <- ordinate(ps.temp, method = "PCoA", distance = "euclidean")

# Plot
temp.temp.week.ord <- plot_ordination(ps.temp, ord) +
  geom_point(
    aes(color = Temperature, shape = Week),
    size = 4,
    alpha = 0.8
  ) +
  stat_ellipse(
    aes(group = Week, color = Week),  # Ellipse by Week
    linetype = "dashed",
    size = 1
  ) +
  scale_color_manual(
    values = c(
      "26" = "blue3", "29" = "darkgreen", "31" = "magenta4", "33" = "red3",
      "1" = "tan4", "3" = "orange", "5" = "brown"  # for Week ellipses
    )
  ) +
  guides(
    shape = guide_legend(override.aes = list(color = "black")), # Week shapes
    color = guide_legend(override.aes = list(shape = 16))        # Temp colors
  ) +
  theme_classic(base_size = 22) +
  theme(
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 18),
    legend.text = element_text(size = 18),
    legend.title = element_text(size = 18)
  )

temp.temp.week.ord
#good enough, will edit key later
```

Set up for Beta Dispersion

-subset seqtab by experiment (sequence table)
```{r}
sum.seqtab <- readRDS("chap4sumseqtab.clr.rds")
field.seqtab <- readRDS("chap4fieldseqtab.clr.rds")
temp.seqtab <- readRDS("chap4tempseqtab.clr.rds")
```

-also samdf is all the metadata, this needs to be read it and subset by experiment as well (already subsetted, so just reading in) (metadata)
```{r}
sum.samdf <- read.csv("chap4sum.samdf.csv")
field.samdf <- read.csv("chap4field.samdf.csv")
temp.samdf <- read.csv("chap4temp.samdf.csv")
```

Make a Distance Matrix: veg-distance calculates pairwise distances between samples (using euclidean distances)
```{r}
sum.dist.all <- vegdist(sum.seqtab, method = "euclidean")
field.dist.all <- vegdist(field.seqtab, method = "euclidean")
temp.dist.all <- vegdist(temp.seqtab, method = "euclidean")
```

Beta Dispersion: 
*looks at spread of points in a multidemensional space
*is beta dispersion differs a lot, PERMANOVA results may be biased

~Summer
```{r}
```


```{r}
sum.bet.all1 <- betadisper(sum.dist.all,sum.samdf$Temperature)
sum.bet.all2 <- betadisper(sum.dist.all,sum.samdf$Week)
sum.bet.all3 <- betadisper(sum.dist.all,sum.samdf$Treatment)
```

*are there true differences?
```{r}
#temp
sum.bet.all1
anova(sum.bet.all1)
```
*no differences in dispersion between temps

```{r}
#week
sum.bet.all2
anova(sum.bet.all2)
TukeyHSD(sum.bet.all2)
```
*diff between:
- weeks 5 & 3
- weeks 5 & 1 


```{r}
#treatment
sum.bet.all3
anova(sum.bet.all3)
```
*no differences in dispersion between treatments

~Field
```{r}
field.bet.all <- betadisper(field.dist.all,field.samdf$Treatment)
field.bet.all
anova(field.bet.all)
```
*no differences in dispersion between  treatments

~Temp
```{r}
#temp
temp.bet.all1 <- betadisper(temp.dist.all,temp.samdf$Temperature)
anova(temp.bet.all1)
```
*no differences in dispersion betweentemperatures


```{r}
#week
temp.bet.all2 <- betadisper(temp.dist.all,temp.samdf$Week)
anova(temp.bet.all2)
TukeyHSD(temp.bet.all2)
```
*diff between:
- weeks 3 & 1
- weeks 5 & 1 

Beta Diversity (PERMANOVAS)


